Equivalence Hypergraphs:\\ Semantics of E-Graphs and their Generalization to Monoidal Theories

### 2023-10-18 sections/categorical_semantics.tex
Fixesing inaccuracies for combinatorial semantics
intro/conclusion

We will assume familiarity with the definition of category,  functor and natural transformation.
\begin{notation*}
We use the following notation: given a catgeory $\mathbb{C}$, we write $ob(\mathbb{C})$ for the set of objects $\mathbb{C}(A,B)$ for hom-sets and $f: A \to B$ for $f \in \mathbb{C}(A,B)$. We write $\textbf{id}_A$ for the identity on $A$ and commonly write $f;g$ for composition in diagram order.  Composition in the usual order is written $g \circ f$. 
We will often omit subscripts on identities and natural transformations where it can be inferred. 
\end{notation*}
\begin{definition}
A \textit{strict symmetric monoidal category (SMC)} $(\mathbb{C}, \otimes, 1)$ is given by a category $\mathbb{C}$ with the following equipment: 
\begin{itemize}
\item a \textit{tensor product} functor $\otimes: \mathbb{C} \times \mathbb{C} \to \mathbb{C}$,
\item a \textit{tensor unit} object $1 \in ob(\mathbb{C})$, 
\item the \textit{symmetry} natural isomorphism: $\sigma: A \otimes B \to B \otimes A$
\item and such that: $A \otimes (B \otimes C) = (A \otimes B) \otimes C$ ,  $1 \otimes A = A$ ,  $A \otimes 1 = A$
\end{itemize}
satisfying certain coherence conditions [MacLane].  A \textit{permutations and products category (PROP)} is a symmetric monoidal category where every object is the n-fold tensor product of some distinguished object $A$. In this case, we will enote the n-fold tensor product of the distinguished object by simply $n$, so that $n \otimes m = n+m$.

@@

For example, the \textit{semilattice signature} consists of a single binary operation $+$.  Given a set $A$, this can be used to freely generate terms according to the following grammar.
\begin{align*}
	\infer{x_1, \ldots, x_n \vdash x_j}{x_i \in A} \qquad \infer{\Gamma \vdash t_1 + t_2}{\Gamma \vdash t_1 && \Gamma \vdash t_2} 
\end{align*}
Such terms may be organized into a Cartesian category, with morphisms given by tuples of terms of the following form,  and composition is given by substitution of terms for free variables.  
\[
	x_1, \ldots, x_n \vdash (t_1, \ldots, t_m) : n \to m\ .
\]
A set of equations can then be given as a set of parallel pairs of morphisms.  
For example, the equations for the theory of semilattices are the following.
\[
	x,y,z \vdash x+(y+z) = (x+y)+z : 3 \to 1\ \qquad  
    x,y \vdash x+y = y+x \qquad 
    x \vdash x+x = x
\]
Now, quotienting the category by the least congruence identifying every such parallel pair, we recover a category whose morphisms are \textit{equivalence classes} of terms modulo the expected equations. 
This gives a categorical presentation of an algebraic theory, known as its \textit{Lawvere category}. 
Importantly,  its category of models (\textit{i.e.}, category of functors to \textbf{Set}) is equivalent to the expected category of models of the algebraic theory in question.

@@

It will often be convenient to present SMTs in terms of string diagrams [which are technically equiv classes of arrows?!] 
\begin{example}
The SMT of \textit{commutative comonoids} is given by the following generators, ${\Delta, !}$, depicted in string diagrammatic form:
\[
    \tikzfig{categorical-semantics/Cartesian-equipment}
\]
together with the following associativity, commutativity and unitality equations, again given in terms of string diagrams. 
\[
    \tikzfig{categorical-semantics/Cartesian-theory}
\]
\end{example}

@@

A \textit{semilattice enriched strict SMC (PROP) functor} $F: (\mathbb{C}, \otimes, 1,+) \to (\mathbb{D}, \otimes, 1,+)$ is given by a strict SMC (PROP) functor on the underlying SMCs (PROPs) which additionally satisfies $F(f+g) = F(f)+F(g)$. 
\end{definition}
Note, in particular,  that semigroups \update{semilattice?} do not require a unit for $+$. {This is important, since, in any commutative monoid enriched category,  if there exists a categorical product, then it must also be a categorical coproduct -- a degeneracy we wish to avoid.}

@@

To aid reasoning, we introduce a new language of string diagrams for semilattice enriched SMCs, using a hierarchical ``box'' structure to capture the binary operation on morphisms.  This will later be used in the translation of equivalence classes from the e-graph to the string diagrammatic setting.

@@

We use string diagrams informally in this paper, but it is intuitively clear that these diagrams are sound and complete with respect to their intended categorical semantics. Indeed, the diagrammatic language itself is not really novel and is similar to \textit{e.g.}, \cite{duncan_generalised_2009}.

>>>>

We will assume familiarity with the definition of category,  functor and natural transformation.
\begin{notation*}
We use the following notation: given a catgeory $\mathbb{C}$, we write $ob(\mathbb{C})$ for the set of objects $\mathbb{C}(A,B)$ for hom-sets and $f: A \to B$ for $f \in \mathbb{C}(A,B)$. We write $\textbf{id}_A$ for the identity on $A$ and commonly write $f;g$ for composition in diagram order.  Composition in the usual order is written $g \circ f$. 
We will often omit subscripts on identities and natural transformations where it can be inferred. 
\end{notation*}
\begin{definition}
A \textit{strict symmetric monoidal category (SMC)} $(\mathbb{C}, \otimes, 1)$ is given by a category $\mathbb{C}$ with the following equipment: 
\begin{itemize}
\item a \textit{tensor product} functor $\otimes: \mathbb{C} \times \mathbb{C} \to \mathbb{C}$,
\item a \textit{tensor unit} object $1 \in ob(\mathbb{C})$, 
\item the \textit{symmetry} natural isomorphism: $\sigma: A \otimes B \to B \otimes A$
\item and such that: $A \otimes (B \otimes C) = (A \otimes B) \otimes C$ ,  $1 \otimes A = A$ ,  $A \otimes 1 = A$
\end{itemize}
satisfying certain coherence conditions 
\cite{maclane}.  A \textit{permutations and products category (PROP)} is a symmetric monoidal category where every object is the n-fold tensor product of some distinguished object $A$. In this case, we will enote the n-fold tensor product of the distinguished object by simply $n$, so that $n \otimes m = n+m$.

@@

For example, the \textit{semilattice signature} consists of a single binary operation $+$.  Given a set $A$, this can be used to freely generate terms according to the following grammar.
\begin{align*}
	\infer{\Gamma \vdash x}{x \in \Gamma} \qquad \infer{\Gamma \vdash a}{a \in A} \qquad \infer{\Gamma \vdash t_1 + t_2}{\Gamma \vdash t_1 && \Gamma \vdash t_2} 
\end{align*}
Such terms may be organized into a Cartesian category, with morphisms given by tuples of terms of the following form,  and composition is given by substitution of terms for free variables.  
\[
	x_1, \ldots, x_n \vdash (t_1, \ldots, t_m) : n \to m\ .
\]
A set of equations can then be given as a set of parallel pairs of morphisms.  
For example, the equations for the theory of semilattices are the following.
\[
	x,y,z \vdash x+(y+z) = (x+y)+z : 3 \to 1\ \qquad  
    x,y \vdash x+y = y+x \qquad 
    x \vdash x+x = x
\]
Now, quotienting the category by the least congruence identifying every such parallel pair, we recover a category whose morphisms are \textit{equivalence classes} of terms modulo the expected equations. 
This gives a categorical presentation of an algebraic theory, known as its \textit{Lawvere category}. 
Importantly,  its category of models (\textit{i.e.}, category of functors to \textbf{Set}) is equivalent to the expected category of models of the algebraic theory in question.

@@

It will often be convenient to give presentations of SMTs in terms of string diagrams.
\begin{example}
The SMT of \textit{commutative comonoids} is given by the following generators, ${\Delta, !}$, depicted in string diagrammatic form:
\[
    \tikzfig{categorical-semantics/Cartesian-equipment}
\]
together with the following associativity, commutativity and unitality equations, again given in terms of string diagrams. 
\[
    \tikzfig{categorical-semantics/Cartesian-theory}
\]
\end{example}

@@

A \textit{semilattice enriched strict SMC (PROP) functor} $F: (\mathbb{C}, \otimes, 1,+) \to (\mathbb{D}, \otimes, 1,+)$ is given by a strict SMC (PROP) functor on the underlying SMCs (PROPs) which additionally satisfies $F(f+g) = F(f)+F(g)$. 
\end{definition}
Note, in particular,  that semilattices do not require a unit for $+$. {This is important, since, in any commutative monoid enriched category,  if there exists a categorical product, then it must also be a categorical coproduct -- a degeneracy we wish to avoid.}

@@

To aid reasoning, we introduce a new language of string diagrams for semilattice enriched SMCs, using a hierarchical ``box'' structure to capture the binary operation on morphisms.  It was already discussed in the introduction how this is  used in the translation of equivalence classes from the e-graph to the string diagrammatic setting.

@@

We use string diagrams informally in this paper, but it is intuitively clear that these diagrams are sound and complete with respect to their intended categorical semantics. Indeed, the diagrammatic language itself is not truly novel and is similar to \textit{e.g.}, \cite{duncan_generalised_2009}.

### 2023-10-18 sections/conclusion.tex
Fixesing inaccuracies for combinatorial semantics
update intro
intro/conclusion

\section{Conclusion and Further Work}

\begin{itemize}
\item implementation for ZX
\item computational linguistics
\item traced categories, hypergraphs for traced comonoids and full e-graphs
\item closed categories, hierarchical hypergraphs and lambda-calculus (example)
\end{itemize}

>>>>

% !TEX root = ../main.tex

\section{Conclusion}

We have described a categorical semantics for e-graphs over algebraic theories in terms of Cartesian categories enriched over a semilattice, and defined a natural generalization to account for monoidal theories in terms of SMCs,  similarly enriched.  Having described a translation of e-graphs into our formalism, we developed a combinatorial representation of morphisms in our free category which factor out the structural SMC equations, but remain sensitive to the complexity-relevant equations induced by the enrichment.  The design of such a representation and its corresponding  rewrite theory proved delicate, and -- together with its proof of completeness -- constitute the main technical contributions of the paper.  Finally, we considered an analogue of e-rewriting in our setting, pushing towards the practical considerations necessary for an implementation of our ideas.  Although we believe our work forms a solid basis for a formal mathematical theory of (acyclic) e-graphs,  there are a number of extensions which we intend to investigate in the future.

\subsection{Future Work and Applications}

We have already alluded to the fact we deal with only \textit{acyclic} e-graphs in this paper.  Nevertheless,  within this setting we present a generalization from Cartesian to (symmetric) monoidal theories,  giving rise to a host of potential new domains of application.  Further,  the string diagrammatic and graph rewriting framework we set up is ameenable to extension, including lifting the acyclicity restriction.

\begin{figure}\label{fig:applications}
\[
	\scalebox{0.75}{
	\tikzfig{combinatorial_semantics/example-e-graphs}
	}
\]
\caption{Example applications of e-hypergraphs.  Left, the ZX-calculus.  Right, the $\lambda$-calculus as represented by string diagrams for Cartesian closed categories.  }
\end{figure}

Symmetric monoidal theories have received significant attention over recent years: the ZX-calculus can be used to characterize quantum processes, with applications to optimization of quantum circuits [cite] ; similar theories also characterize signal flow diagrams 
\cite{baldan_categorical_2014}, digital circuits 
\cite{ghica_operational_2021}, as well as having application to computational linguistics \cite{wazni_quantum_2022,coecke_lambek_2013}.  In future work, we hope to apply our generalization of e-graphs to optimization problems in these domains (e.g. Figure \ref{fig:applications})

The theories described above often have more structure than simply being symmetric monoidal: for example, they may contain comonoids, Frobenius algebras or traced structure (as in the case of e-graphs with cycles),  and this structure can be exploited to give more natural (in the sense of factoring out more equations) combinatorial representations.  In the above cases, such representations and their DPO-rewriting theory is well-studied 
\cite{ghica_rewriting_2023} % traced comonoid
[cite, cite, cite].  We consider that our development of hierarchical "e-box" structure within string diagrams and their associated (e-)hypergraphs ( categorically,  semilattice enrichment) is essentially orthogonal to these developments, and so ought to be easily combined to give notions of e-hypergraphs specifically suited to the optimization of these particular monoidal theories.

A particularly interesting extension that we intend to consider is the combination of our "e-structure" with (Cartesian) \textit{closed} categories, allowing the natural combination of e-graph optimization techniques with the $\lambda$-calculus.  The corresponding combinatorial representation would build on the recent work \cite{ghica_hierarchical_2023, alvarez-picallo_rewriting_2022, alvarez-picallo_functorial_2021}. % hierarchical

In Figure \ref{fig:let-calculus} we give an example term calculus corresponding to our categorical semantics.  Although undeveloped here, the first author intends to develop this idea, extending the calculus of \cite{heijltjes_functional_2023,barrett_functional_2023} into a natural term language for such string diagrams. 
We also mention that our developments would also naturally combine with recent work on string diagrams for \textit{non-strict} symmetric monoidal categories \cite{wilson_string_2023}.  % non-strict categories
Finally, we intend to develop implementations of the theory developed here,  with an ultimate aim of providing a tool for optimization of monoidal theories and assessing its practicality on real-world problems, such as those considered in the various domains mentioned above.

### 2023-10-18 sections/introduction.tex
intro/conclusion

\cite{heijltjes_functional_2023}
\cite{barrett_functional_2023}
\cite{griggio_proceedings_2022}
\cite{flatt_small_2022}
\cite{ghica_operational_2021}
 % digital circuits
\cite{ghica_hierarchical_2023}
\cite{ghica_rewriting_2023} % traced comonoid
\cite{wilson_string_2023} % non-strict categories

\cite{zhang_relational_2022} % relational ematching 
\cite{alvarez-picallo_rewriting_2022} % hierarchical
\cite{alvarez-picallo_functorial_2021} % hierarchical

\cite{bonchi_tape_nodate} % tape diagrams
\cite{baldan_categorical_2014} % signal flow
\cite{dpo}
\cite{maclane}
\cite{singher2023colored}

@@

A recently proposed,  alternative solution to this problem is \textit{equality saturation} 
\cite{10.1145/1594834.1480915}: instead of maintaining a \textit{single},  putatively optimized program which is rewritten \textit{destructively} at each step, we can maintain a \textit{set} of equivalent programs, where each rewrite step \textit{non-destructively} grows the set.  Upon reaching a fixed point (\textit{saturation}),  a \textit{globally} optimal program can then be extracted.% from the set,  which now represents all equivalent ways to express the program with respect to the given rewrites.

The use of \textit{equality graphs (e-graphs)} \cite{EggPaper} -- a generalization of the typical directed acylcic graph (DAG) representation of terms to include equivalence classes -- to compactly represent the set of equivalent programs is what makes the technique tractable in practise, where the naive approach is clearly unfeasible.

\update{Move focus from semantics of e-graphs to their extension to monoidal theories?}
While equality saturation is a state-of-the-art optimization technique, the formal mathematical theory of e-graphs is perhaps underdeveloped. In this paper, we contribute a categorical understanding of e-graphs and their rewriting. Considering programs as represented by terms of an arbitrary algebraic theory, we demonstrate a correspondence between e-graphs and (string diagrams for) simple \textit{semilattice-enriched} Cartesian categories, where the semilattice \textit{join} allows us to consider formal sets of terms (string diagrams, morphisms).

However, we do not develop the categorical perspective purely for its own sake: rephrasing the formalism of e-graphs in terms of algebraic theories and Cartesian categories suggests a natural generalization to monoidal theories. Therefore, our theoretical development leads to new potential domains of application, including optimziation of string diagrams representing digital \cite{ghica_compositional_2023} and quantum circuits
\cite{coecke_interacting_2011,ZX} and potentially to natural language \cite{wazni_quantum_2022,coecke_lambek_2013}.  We also suggest how our approach will allow future  generalization to include rewriting for the $\lambda$-calculus 
\cite{koehler2022sketchguided} % latest egraph.

Towards implementation, we 
additionally provide a combinatorial representation of our enriched string diagrams in terms of (hierarchical)  hypergraphs.  We give a specification of hypergraph rewriting in this novel setting via a suitable extension of the double pushout (DPO) rewriting framework 
\cite{bonchi_string_2022-1,bonchi_string_2022-2,bonchi_string_2022},  thus also formalizing the rewriting theory of standard e-graphs.

\subsection*{E-graphs}

@@

\subsection*{Semantics via Semilattice Enriched Categories}

One can easily see the correspondence between (acyclic) e-graphs and morphisms of a suitably enriched Cartesian category by considering string diagrams 
\cite{noauthor_09083347_nodate,joyal_geometry_1991, mellies_functorial_2006}. In Figure \ref{fig:egraph-strings} we have extended the usual string diagrammatic notation for symmetric monoidal categories with an additional generator which has the  typing rule below, left.
The ability to take formal sums of morphisms is used to model the equivalence class structure of e-graphs.  
\[
\infer{\phi_1 + \ldots + \phi_n: A \to B}{\{\phi_i: A \to B\}_{i \in \{1,\ldots,n\}}}
\qquad
\infer{\Gamma \vdash \{t_1, \ldots, t_n\}: A}{\{\Gamma \vdash t_i: A\}_{i \in \{1, \ldots, n\}}}
\]
To aid understanding, in Figure \ref{fig:let-calculus} we give an example of a representation of the string diagrams in Figure \ref{fig:e-graph-example} using an informal let-calculus, which is standard except for its augmentation with the typing rule above, right, corresponding to semilattice enrichment. The new construct forms a set of terms and satisfies the following additional congruence: $K\{\{t_1, \ldots, t_n\}\} = \{K\{t_1\}, \ldots, K\{t_n\}\}$.

@@

Note how the natural level of generality for our string diagrams is not Cartesian, but monoidal: the generators of a Cartesian category are required in order to embed e-graphs, but are orthogonal to the addition of formal sums. Thus we generalize e-graphs from algebraic to monoidal theories.

\subsection*{Combinatorial Representation of Enriched String Diagrams}

In order to \textit{implement} our generalized version of e-graphs, we will  perform rewriting not on algebraic terms, but on combinatorial representations of the corresponding string diagrams.  String diagrams can be considered equivalently as either topological objects (\textit{i.e.}, taken modulo "connectivity") or as a 2-dimensional syntax quotiented by the equations of an SMC. For example, we have the following equivalences between diagrams. 
%\[(f_1 \otimes f_2) ; (g_1 \otimes g_2) = (f_1 ; g_1) \otimes (f_2 ; g_2)\] 
\[
	\tikzfig{categorical-semantics/interchange}
\]

Unfortunately, this makes string diagrams relatively unamenable to efficient implementation as a data structure, due to the difficulty of calculating the quotient.

Nevertheless, there is a body of work dealing with this issue by representing string diagrams as combinatorial objects -- namely, (appropriately generalized) hypergraphs 
\cite{bonchi_string_2022-1,bonchi_string_2022-2,bonchi_string_2022}.  Here, the generators $c_i$ become vertices which are connected via hyper-edges.  Thus the expected quotient becomes simply hypergraph isomorphism.  With appropriate restrictions on the form these hypergraphs can take, this approach can be used to \textit{characterize} the free symmetric monoidal category generated by some $c_i$.

We are interested in not just the free SMC over a set of generators, but rather in \textit{symmetric monoidal theories} which involve considering extra equations between morphisms: for instance, the rewrite rules $(a)-(e)$ of Figure \ref{fig:e-graph-example}. These can be seen as equations between string diagrams, and thus between their hypergraph representations.  Because the generating equations can be applied in any context, we are lead to the notion of \textit{(hyper)graph rewriting}: given an equation $l=r$, we require some way to identify (a sub-hypergraph corresponding to) $l$ in any hypergraph $G$ and replace it with (a sub-hypergraph corresponding to) $r$.
The standard approach to this is known as\textit{double pushout (DPO) rewriting}.

Our application requires a generalization of these concepts from hypergraphs to \textit{e-hypergraphs}, which we define as a hypergraph with two additional relations denoting the hierarchical structure introduced by "e-boxes" ( \textit{i.e.,} the generator for semilattice enrichment) and the separation of the components of each e-box.
Giving the appropriate definitions is slightly delicate matter, and constitutes the main body of the paper.

\subsection*{Soundness and Completeness}
The main technical result of the paper is a proof soundness and completeness of the combinatorial representation for semilattice enriched SMCs, extending the results of \cite{bonchi_string_2022-2} for plain SMCs. While the structural equalities of SMCs are factored out in the representation,  important structural equalities arising from the enrichment (see the equations of Figure \ref{fig:string-equations}) are not, and should not be: they represent the un/sharing of subdiagrams with respect to the e-box structure,  which is precisely what allows for the compact representation of equivalence classes.  Instead, we consider DPO-rewrites implementing both the structural equalities for enrichment (which involve the e-box structure) and the equalities arising from the generating symmetric monoidal theory (which do not).

Quotienting e-hypergraphs by the induced equivalence gives rise to a sound and complete model of semilattice enriched categories.  That is, we exhibit a translation functor $\llbracket-\rrbracket$ from the free semilattice enriched SMC over a symmetric monoidal theory to our (quotiented) category of e-hypergraphs such that 
morphisms $f = g$ if and only if there exists a sequence of DPO-rewrites (each induced by a structural equality or the theory) between their translations: $\llbracket f \rrbracket \rightsquigarrow \llbracket_{DPO} g \rrbracket$.

\subsection*{E-matching and E-rewriting}

Having defined e-hypergraphs and shown them to correctly model rewriting string diagrams for SMCs, equipped with a notion of "equivalence class", we note that the naive implementation of DPO-rewriting is inefficient: to find a redex within an e-hypergraph involves finding a structurally equivalent e-hypergraph which contains the redex as a subgraph. In general, this involves unsharing the e-box structure before a redex becomes available. Thus, we define a notion of e-matching for e-hypergraphs: that is, to find redexes working modulo the e-structure. In particular, we wish to locate the smallest subgraph $G' \subset G$ "containing" the redex $L$. Given this, the analogue of e-rewriting $L \to R$ is simple to define, due to its non-destructive nature: we rewrite $G' \to G' + R$ in $G$.

[Example diagrams]

\subsection*{Future Applications}

Example diagrams for circuits, trace (loop unrolling, PEGs), LC, ZX here.

\subsection{Related Work}
Q: Related work - at end? \\

>>>>

@@

A recently proposed,  alternative solution to this problem is \textit{equality saturation} 
\cite{10.1145/1594834.1480915}: instead of maintaining a \textit{single},  putatively optimized program which is rewritten \textit{destructively} at each step, we can maintain a \textit{set} of equivalent programs, where each rewrite step \textit{non-destructively} grows the set.  Upon reaching a fixed point ({saturation}),  a \textit{globally} optimal program can then be extracted.% from the set,  which now represents all equivalent ways to express the program with respect to the given rewrites.  
The use of \textit{equality graphs (e-graphs)} \cite{EggPaper} -- a generalization of the typical directed acylcic graph (DAG) representation of terms to include equivalence classes -- to compactly represent the set of equivalent programs is what makes the technique tractable in practise, where the naive approach is clearly unfeasible.

While equality saturation is a state-of-the-art optimization technique, the formal mathematical theory of e-graphs is perhaps underdeveloped. In this paper, we contribute a categorical understanding of e-graphs and their rewriting. Considering programs as represented by terms of an arbitrary algebraic theory, we demonstrate a correspondence between e-graphs and (string diagrams for) simple \textit{semilattice-enriched} Cartesian categories, where the semilattice \textit{join} allows us to consider formal sets of terms (string diagrams, morphisms).

However, we do not develop the categorical perspective purely for its own sake: rephrasing the formalism of e-graphs in terms of algebraic theories and Cartesian categories suggests a natural generalization to monoidal theories. Therefore, our theoretical development leads to new potential domains of application, including optimziation of string diagrams representing digital \cite{ghica_compositional_2023} and quantum circuits
\cite{coecke_interacting_2011,ZX} and potentially to computational linguistics \cite{wazni_quantum_2022,coecke_lambek_2013}.  We also suggest how our approach will allow future  generalization to include rewriting for the $\lambda$-calculus,  in a novel approach \textit{cf.} \textit{e.g.,}
\cite{koehler2022sketchguided}. % latest egraph.

With consideration of future implementations, we 
additionally provide a combinatorial representation of our enriched string diagrams in terms of (hierarchical)  hypergraphs, which we call \textit{e-hypergraphs}.  We give a specification of e-hypergraph rewriting via a suitable extension of the double pushout (DPO) rewriting framework 
\cite{dpo, bonchi_string_2022-1,bonchi_string_2022-2,bonchi_string_2022}, and prove the representation sound and complete with respect to our categorical semantics.  As a corollary, we also formalize the rewriting theory of standard e-graphs in this way.  Finally, we touch on the question of efficient implementation of rewriting for e-hypergraphs.

\subsection{E-graphs}

@@

\subsection{Semantics via Semilattice Enriched Categories}

Recalling the correspondence between DAG representations of algebraic terms and the corresponding string diagrammatic representation of morphisms of a suitable Cartesian category, 
one can easily see the correspondence between (acyclic) e-graphs and morphisms of a suitably enriched Cartesian category by considering string diagrams 
\cite{noauthor_09083347_nodate,joyal_geometry_1991, mellies_functorial_2006}. In Figure \ref{fig:egraph-strings} we have extended the usual string diagrammatic notation for symmetric monoidal categories with an additional generator which has the  typing rule below, left, corresponding to taking formal sums of morphisms.
The ability to take formal sums is used to model the equivalence class structure of e-graphs.  
\[
\infer{\phi_1 + \ldots + \phi_n: A \to B}{\{\phi_i: A \to B\}_{i \in \{1,\ldots,n\}}}
\qquad
\infer{\Gamma \vdash \{t_1, \ldots, t_n\}: A}{\{\Gamma \vdash t_i: A\}_{i \in \{1, \ldots, n\}}}
\]
To aid understanding, in Figure \ref{fig:let-calculus} we give an example of a representation of the string diagrams in Figure \ref{fig:e-graph-example} using an informal let-calculus, which is standard except for its augmentation with the typing rule above, right, corresponding to semilattice enrichment. The new construct forms a set of terms and satisfies the following additional congruence: $K\{\{t_1, \ldots, t_n\}\} = \{K\{t_1\}, \ldots, K\{t_n\}\}$.

@@

Note how the natural level of generality for our string diagrams is not Cartesian, but monoidal: the generators of a Cartesian category are required in order to embed e-graphs, but are orthogonal to the addition of formal sums. Thus we generalize e-graphs from algebraic to monoidal theories, giving rise to a  host of new potential applications, which we outline in the conclusion of this paper.

\subsection{Combinatorial Representation of Enriched String Diagrams}

In order to \textit{implement} our generalized version of e-graphs, we will  perform rewriting not on terms representing morphisms of a category, but on combinatorial representations of the corresponding string diagrams.  String diagrams can be considered equivalently as either topological objects (\textit{i.e.}, taken modulo "connectivity") or as a 2-dimensional syntax quotiented by the equations of an SMC. For example, we have the following equivalences between diagrams. 
%\[(f_1 \otimes f_2) ; (g_1 \otimes g_2) = (f_1 ; g_1) \otimes (f_2 ; g_2)\] 
\[
	\tikzfig{categorical-semantics/interchange}
\]

Unfortunately, this makes string diagrams relatively unamenable to efficient implementation as a data structure, due to the difficulty of calculating the quotient. 
Nevertheless, there is a body of work dealing with this issue by representing string diagrams as combinatorial objects -- namely, (appropriately generalized) hypergraphs 
\cite{bonchi_string_2022-1,bonchi_string_2022-2,bonchi_string_2022}.  Here, the generators $c_i$ become vertices which are connected via hyper-edges.  Thus the expected quotient becomes simply hypergraph isomorphism.  With appropriate restrictions on the form these hypergraphs can take, this approach can be used to \textit{characterize} the free symmetric monoidal category generated by some $c_i$.

We are interested in not just the free SMC over a set of generators, but rather in \textit{(symmetric) monoidal theories} which involve considering extra equations between morphisms: for instance, the rewrite rules $(a)-(e)$ of Figure \ref{fig:e-graph-example}. These can be seen as equations between string diagrams, and thus between their hypergraph representations.  Because the generating equations can be applied in any context, we are lead to the notion of \textit{(hyper)graph rewriting}: given an equation $l=r$, we require some way to identify (a sub-hypergraph corresponding to) $l$ in any hypergraph $G$ and replace it with (a sub-hypergraph corresponding to) $r$.
We make use of the standard approach to giving specifications of such graph rewriting, known as \textit{double pushout (DPO) rewriting}.

Our application requires a generalization of these concepts from hypergraphs to \textit{e-hypergraphs}, which we define as a hypergraph with two additional relations denoting the hierarchical structure introduced by \textit{e-boxes} (\textit{i.e.,} the generator for semilattice enrichment) and the separation of the components of each e-box.
Giving the appropriate definitions is slightly delicate matter, and constitutes the main body of the paper.

\subsection{Soundness and Completeness}
The main technical result of the paper is a proof soundness and completeness of the combinatorial representation for semilattice enriched SMCs, extending the results of \cite{bonchi_string_2022-2} for plain SMCs. While the structural equalities of SMCs are factored out in the representation,  important structural equalities arising from the enrichment (see the equations of Figure \ref{fig:string-equations}) are not, and should not be: they represent the un/sharing of subdiagrams with respect to the e-box structure,  which is precisely what allows for the compact representation of equivalence classes.  Instead, we consider DPO-rewrites implementing both the structural equalities for enrichment (which involve the e-box structure) and the equalities arising from the generating monoidal theory (which do not).

Quotienting e-hypergraphs by the induced equivalence gives rise to a sound and complete model of semilattice enriched SMCs.  That is, we exhibit a translation functor $\llbracket-\rrbracket$ from the free semilattice enriched SMC over a symmetric monoidal theory to our (quotiented) category of e-hypergraphs such that 
morphisms $f = g$ if and only if there exists a sequence of DPO-rewrites (each induced by a structural equality or the monoidal theory) between their translations: $\llbracket f \rrbracket \rightsquigarrow_{DPO} \llbracket g \rrbracket$.

\subsection{E-matching and E-rewriting}

Having defined e-hypergraphs and shown them to correctly model the rewriting of string diagrams for semilattice enriched SMCs, we note that the naive implementation of DPO-rewriting is inefficient: to find a redex within an e-hypergraph involves finding a structurally equivalent e-hypergraph which contains the redex as a subgraph. In general, this involves unsharing the e-box structure before a redex becomes available.  Thus, we define a notion of \textit{e-matching} for e-hypergraphs: that is, to find redexes working modulo the e-box structure. In particular, we wish to locate the smallest subgraph $G' \subseteq G$ "containing" (in an appropriate sense) the redex $L$. Given this,  \textit{e-rewriting} $L \to R$, intended to achieve equality saturation, is simple to define due to its non-destructive nature: we rewrite $G' \to G' + R$ in $G$.  
\begin{figure}
\[
	\tikzfig{combinatorial_semantics/example-rewriting}
\]
\caption{Example application of e-rewriting for rewrite $(x*y)/z \to x*(y/z)$: find an appropriate minimal sub-e-hypergraph containing the redex $(x*y)/z$ and non-destructively add the reduct $x*(y/z)$.  REDO THIS}
\end{figure}

\subsection{Related Work}

Although e-graphs were first developed in the 80's \cite{nelson1980techniques}, there has recently been an explosion of interest in them for the purpose of building program optimizers and synthesizers, especially due to recent work on the equality saturation technique \cite{10.1145/1594834.1480915, griggio_proceedings_2022, EggPaper,flatt_small_2022}.  This includes interest in various extensions of the e-graph formalism, for example to account for conditional rewriting \cite{singher2023colored},  to reason about rewriting for the $\lambda$-calculus \cite{koehler2022sketchguided},  and to combine techniques of e-graphs and database queries ("relational e-matching") \cite{zhang_relational_2022}.  We are hopeful that our novel perspective on e-graphs can be extended to give foundations to these investigations.  Some further avenues for investigation, including potential applications, are given in the conclusion of this paper.

Our use of string diagrams is also central to the intuition behind our work.  The string diagrams we work with are left essentially as informal reasoning devices, however there exists a breadth of work on similar diagrammatic formalisms for similar classes of categories.  We mention here work on hierarchical string diagrams \cite{ghica_hierarchical_2023},  functorial boxes \cite{mellies_functorial_2006},  proof nets for compact closed categories with biproducts \cite{duncan_generalised_2009}, and recent work on tape diagrams for rig categories with biproducts \cite{bonchi_tape_nodate}.
